Modularly Composing Typed Language Fragments

### 2014-11-02 att-pldi15/att-pldi15.tex
Finished draft of PLDI15 except some citations

\maketitle
\begin{abstract}
Researchers often describe typed programming languages as fragments or simple calculi, leaving to language designers the task of composing these to form complete languages.  %comparable to those available for library-based embeddings packaged using a modern module system. 
This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
As the design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries. 
%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.

@@

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant of record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of associated term-level operators (e.g. record projection) using functions written in a static language where types and translations are values. The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension. Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing abstraction barriers around tycons using type abstraction, the same principle that underlies ML-style modules.
\end{abstract}

@@

Universality properties (i.e. ``Turing-completeness'') can be invoked to guarantee that an embedding that preserves a desirable fragment's dynamic semantics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed).

These problems do not merit concern when a fragment is expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We describe  related work in Sec. \ref{prior-work} and conclude with a brief discussion of expressiveness in Sec. \ref{discussion}.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds will be needed: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

@@

The rule (k-otherty) governs types constructed by an ``other'' tycon. These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec. \ref{metatheory}. The index of such a type must pair a natural number with a type translation of kind $\kity$, discussed in Sec. \ref{sec:type-translations}.

\paragraph{Examples}
As our first example, consider the user-defined tycon $\tcvar{rstr}$. The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expression patterns and we assume is defined as an inductive sum kind in the usual way. Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision. Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion). We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels. The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$.

In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$. Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.

\subsection{Static Type Case Analysis}
A type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. 
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}
The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (these always take the default branch).

Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}. %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-5px}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

@@

\subsection{Type Translations}\label{sec:type-translations}
Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$. A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).

@@

The kinding rules (and dynamics) for these shared forms simply proceed recursively, e.g.
\begin{mathpar}
\small
\inferrule[k-ity-prod]{
    \sofkX{\sqity{\qity_1}}{\kity}\\
    \sofkX{\sqity{\qity_2}}{\kity}
}{
    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
}
\end{mathpar}
%The operational semantics for shared forms are also straightforwardly recursive (see supplemental material).

The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to the translation of another type $\st$: \begin{mathpar}\small
\inferrule[k-ity-unquote]{
    \sofkX{\st}{\kity}   
}{
    \sofkX{\sqity{\qtuq{\st}}}{\kity}
}

@@

Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type). For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$. The term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity. Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e. a list constructed from the row labels is the term index and the corresponding row values are the arguments. In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms. Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.

@@

The argument interface list, called $\stx{args}$ in the rules, is simply a static list of kind $\kargs$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$. It is generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is given as the length of the argument list, written $|\es|=n$ (see supplement).

>>>>

\maketitle
\begin{abstract}
Researchers often describe type systems as fragments or simple calculi, leaving to language designers the task of composing these to form complete programming  languages.  %comparable to those available for library-based embeddings packaged using a modern module system. 
This is not a systematic process: metatheoretic results must be established anew for each composition, guided only notionally by metatheorems derived for simpler systems.
As the language design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.%, so that important metatheoretic results need only be established locally and, ultimately, so that language extensions can be treated like libraries. 
%,  Sometimes this fails because certain language features do not ``play well'' together, but precisely characterizing which is elusive given these practices.

@@

In this paper, we begin from first principles with a core calculus, @$\lambda$, specified like many full-scale languages: as a bidirectionally typed translation semantics. Only the $\rightarrow$ type constructor (tycon) is built in; all other external tycons (we show a variant on record types and constrained string types) are defined by extending a \emph{tycon context}. Each tycon defines the semantics of its associated term-level operators (e.g. record projection) using functions written in a static language where types and translations are values. The semantics provide  strong metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that all \emph{tycon  invariants} will be conserved under extension. Mechanized proofs are not needed: problems are caught during typechecking by lifting typed compilation techniques into the semantics and enforcing barriers around tycons using type abstraction, the same principle that underlies reasoning with ML-style modules.
\end{abstract}

@@

Universality properties (e.g. ``Turing-completeness'') can usually be invoked to guarantee that an embedding that preserves a desirable fragment's dynamics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a fragment like this is needed, a new \emph{dialect} of a language must be constructed. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed between every pair of dialects).

These problems do not arise for a fragment that can be expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about language fragments. We conclude by discussing  related and future work in Sec. \ref{prior-work}.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds will be needed: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$, indexed by a natural number $m$ to allow arbitrarily many such tycons. The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\sval{\st}{\cdot;\emptyset;\Phi}$.

@@

The rule (k-otherty) governs types constructed by an ``other'' tycon. These will serve only as technical devices to stand in for types other than those in a given tycon context in Sec. \ref{metatheory}. The index of such a type must pair a natural number with a static value of kind $\kity$, discussed in Sec. \ref{sec:type-translations}.

\paragraph{Examples}
Our first example of a user-defined tycon is $\tcvar{rstr}$. It has index kind $\krx$, which classifies static regular expression patterns (defined as an inductive sum kind in the usual way). Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision; recent work has specified how to define type-specific (or here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

Our second example is the tycon $\tcvar{lprod}$, which will define a variant of labeled product type (labeled products are like record types, but maintain a row  ordering; record types are also definable in a manner discussed in the supplement, but maintaining an ordering simplifies our discussion). We choose the index kind of $\tcvar{lprod}$ to be $\klist{\klbl \times \kty}$, where list kinds are defined as inductive sums in the usual way, and $\klbl$ classifies static representations of row labels. The tycon context containing only $\tcvar{lprod}$'s definition is $\small\Phi_\text{lprod} := \tcdef{\tcvar{lprod}}{\tcsig{\klist{\klbl \times \kty}}{\chi_\text{lprod}}}{\theta_\text{lprod}}$. In the tycon context containing both tycon definitions, $\Phi_\text{rstr}\Phi_\text{lprod}$, we can define a labeled product type classifying conference papers, $\stx{paper} := \sty{\tcvar{lprod}}{\{\conclbl{title} : \stx{title}, \conclbl{conf} : \stx{conf}\}}$. Note that $\istype{\stx{paper}}{\Phi_\text{rstr}\Phi_\text{lprod}}$ and we again use kind-specific syntax, in this case for $\klbl$ and $\klist{\klbl \times \kty}$.

\subsection{Static Type Case Analysis}
Types in @$\lambda$ can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context \cite{conf/ppdp/LohH06}. Consistent with this view, a type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. 
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}
The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (such types always take the default branch).

 %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-5px}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

@@

\subsection{Type Translations}\label{sec:type-translations}
Recall that every type $\st$ must have a type translation, $\tau$. Each tycon in the tycon context {computes} translations for the types it constructs as a function of each type's index by specifying a \emph{translation schema} in the {tycon structure}, $\theta$. A tycon with index kind $\ktyidx$ defines a translation schema of kind $\ktyidx \rightarrow \kity$, checked by (tcc-ext).

@@

The kinding and dynamics for these shared forms simply proceed recursively, e.g.
\begin{mathpar}
\small
\inferrule[k-ity-prod]{
    \sofkX{\sqity{\qity_1}}{\kity}\\
    \sofkX{\sqity{\qity_2}}{\kity}
}{
    \sofkX{\sqity{\qity_1 \times \qity_2}}{\kity}
}
\end{mathpar}
%The operational semantics for shared forms are also straightforwardly recursive (see supplemental material).

The syntax for translational internal types additionally includes an ``unquote'' form,  $\qtuq{\st}$, so that they can be constructed compositionally, as well as a form, $\srep{\st}$, that allows one type's translation to refer to another: \begin{mathpar}\small
\inferrule[k-ity-unquote]{
    \sofkX{\st}{\kity}   
}{
    \sofkX{\sqity{\qtuq{\st}}}{\kity}
}

@@

Before discussing rules (ana-intro) and (ana-intro-other), we note that we can recover a variety of standard concrete introductory forms by a purely syntactic desugaring to this abstract form (and thus allow their use at more than one type). For example, for regular strings we can use the string literal form, $\texttt{"s"}$, which desugars to $\eintro{\texttt{"s"}_\text{SL}}{\cdot}$, i.e. the term index is the corresponding static value of kind $\keyw{Str}$, indicated by a subscript for clarity. Similarly, for labeled products, records, objects and so on, we can define a generalized labeled collection form, $\{\mathtt{lbl}_1=e_1, \ldots, \mathtt{lbl}_n=e_n\}$, that desugars to $\eintro{\texttt{[}\mathtt{lbl}_1, \ldots, \mathtt{lbl}_n\texttt{]}}{e_1; \ldots; e_n}$, i.e. a list constructed from the row labels is the term index and the corresponding row values are the arguments. In both cases, the term index captures  static portions of the concrete form and the arguments capture all external sub-terms. Additional desugarings are shown in the supplement and a technique based on \cite{TSLs} could be introduced to allow tycon providers to define more such desugarings  composably.

@@

The argument interface list generated by the judgement $\keyw{args}(n)=\stx{args}$, where $n$ is the length of the argument list, written $|\es|=n$, is then simply a static value of kind $\kargs$ of length $n$ where the $i$th entry is $(\slam{\kty}{\svar{ty}}{\sana{i}{\svar{ty}}}, \slam{\kunit}{\_}{\ssyn{i}})$.
